Temperature-based MHC class-I multimer peptide exchange for human HLA-A, B and C

Version 1

Abstract

Abstract

T cell recognition of specific antigens presented by major histocompatibility complexes class-I (MHC-I) can play an important role during immune responses against pathogens and cancer cells. Detection of T cell immunity is based on assessing the presence of antigen-specific cytotoxic CD8+ T cells using MHC class-I (MHC-I) multimer technology. Previously we have designed conditional peptides for HLA-A*02:01, H-2Kb and HLA-E that form stable peptide-MHC-I-complexes at low temperatures and dissociate when exposed to a defined elevated temperature. The resulting conditional MHC-I complex can easily and without additional handling be exchanged with a peptide of interest, allowing to exchange peptides in a ready-to-use multimer and a high-throughput manner. Here we present data that this peptide-exchange technology is a general applicable, ready-to-use and fast approach to load many different peptides in MHC-I multimers for alleles of the HLA-A, HLA-B and HLA-C loci. We describe the development of conditional peptides for HLA-A*03:01, HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02 that only form stable peptide-MHC-I complexes at low temperatures, allowing peptide exchange at higher defined temperature. We document the ease and flexibility of this technology by monitoring CD8+ T cell responses to virus-specific peptide-MHC complexes in patients.

Introduction

An essential component of cellular immunity is mediated through the T Cell Receptor (TCR) of CD8+ T cell that can bind MHC-I complexes. The MHC-I complex consist of three components: a polymorphic heavy chain, the β2-microglobulin (β2M) protein, and an 8–10-amino-acid peptide ligand[1]. Peptide selection occurs through docking of the peptide C-terminus in MHC-I and occupation of the pockets in MHC-I followed by a conformational change that preferably traps the peptides with highest MHC-binding affinity. As a result, peptides with low MHC-binding affinity are replaced by peptides with a high MHC-binding affinity, a phenomenon used in this work [2]. The peptide ligand is an essential part of the MHC-I complex; without it, the complex cannot remain stable and dissociate[3]. Pathogens, such as viruses, can infect human cells and a primary infection often leads to disease and immunity. Recurrent infections typically do not cause severe disease and are cleared, in part by T cells recognizing viral peptides presented by MHC-I, as recently demonstrated for SARS-CoV-2[4, 5]. The emergence of viral variants can potentially give rise to a partial change in immunogenic viral peptides thereby affecting disease progression. Moreover, immune-compromised patients who lack proper B cell immunity are fully dependent on T cell-mediated immunity. Therefore, monitoring antigen-specific T cells is crucial for tracking the impact of recurrent infections and determining whether therapeutic intervention is required.

The detection and visualization of antigen-specific T cells was first achieved by using peptide-loaded MHC-I multimers with a fluorochrome[6]. To generate such labelled multimers, soluble MHC-I monomers complexed with a peptide are multimerized through biotinylating and fluorochrome-conjugated streptavidin. This process is time-consuming and involves multiple steps for each individual peptide-MHC-I complex. Several technologies have since been developed to allow the parallel production of multiple MHC-I multimers, facilitating high-throughput analysis. For instance, dipeptides have been used as catalysts for peptide exchange[7], and UV-sensitive conditional MHC ligands have been a widely adopted approach, in which a photocleavable peptide is cleaved upon UV exposure. This results in an unloaded MHC-I monomer, which can subsequently be loaded with an high-affinity peptide and multimerized[8, 9]. However, UV exposure can cause photodamage to the peptide MHC-I complex, that leads to sample evaporation, and is incompatible with fluorescently labeled multimers, requiring multimerization to be performed after the peptide exchange[10]. Additionally, more recent methods involve empty MHC-I stabilized by the introduction of a disulfide bond, which can be loaded with the desired a peptide of choice. It is currently unknown how these mutations and structural alterations affect the peptide binding repertoire of these disulfide-stabilized MHC-I molecules[11, 12].

One major limitation of current MHC-I multimer-based technologies is the complex, multistep protocols to obtain ready-to-use peptide MHC-I complexes, without inducing mutations in the HLA class I heavy chain or B2M. This limits the broad clinical usage of MHC-I multimer technology. To address this, we previously developed a temperature-based peptide exchangeable MHC-I multimer technology for human HLA-A*02:01, HLA-E and murine H2Kb, which simplifies and accelerates peptide exchange on MHC-I multimers[13-15]. This technology allows the generation of multimers within hours with ease, simply by incubating selected peptides at a defined temperature with a single standardized batch of fluorescently labeled MHC-I multimers [16]. The MHC-I multimers are multimerized and fluorescently labeled prior to peptide exchange, reducing handling for the end user. In this study, we aimed to determine whether this technology could be developed into a robust and broadly applicable system for analyzing CD8+ T cell immunity. To this end, we developed temperature-based peptide-exchange MHC-I multimers for human MHC gene product HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02, representing all major MHC-I gene families.

Results

Design of temperature-exchangeable template peptide

We set out to extend our toolbox of HLA alleles that allow temperature-based template peptide exchange using multimerized, ready-to-use MHC-I complexes. Our goal was to design peptides that form stable complexes with MHC-I at low temperature and are unstable at higher temperatures, allowing replacement of the template peptide by a desired high-affinity peptide. The peptide off-rate is the primary determinant for the MHC-I stability[17]. For HLA–A*03:01, we designed peptides based on a Hepatitis C virus epitope (HCV43-51; RLGVRATRK) (dissociation constant [Kd] 11.7 nM)[18] and modified the residues to decrease the dissociation constant[19]. HLA–A*03:01 complexes were produced with wild-type or modified peptide, and thermal stability was assessed at 50°C using analytical size exclusion high-performance liquid chromatography (HPLC). With the modified peptide, RGAVRATRR (Kd 7098.44 nM), the complex was unstable at 50°C, resulting in denaturation, as shown by the absence of a MHC-I peak in a size-exclusion HPLC (Figure 1 & Supplemental Figure 1). In contrast, the presence of the high-affinity epitope (RLGVRATRK, 11.7 nM) produced a distinct peak, demonstrating that the HLA-A*03:01 complex could be rescued from unfolding. A similar approach was applied to HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02. For each allele, we designed peptides for each respective allele and modified the residues to decrease the affinity (Table 1)[18, 20, 21]. HLA complexes with the modified peptides were produced, and their thermal stability was evaluated at 50°C using analytical size-exclusion HPLC. For each allele, we identified a thermal template peptide that was unstable at 50°C, which could be rescued by the addition of a high-affinity peptide (Figure 1 & Supplemental Figure 1). We proceeded with peptides that exhibited thermo-instability at 50°C that showed a rescued HLA complex in the presence of the high-affinity peptide.

Thermal stability of MHC-I monomers with template peptide.

Thermal stability of HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, HLA-C*07:01 and HLA-C*07:02 (chromatograms) complexed with template peptide at 50°C, assessed using analytical size exclusion HPLC. Peptide–MHC I monomers were incubated with indicated peptides at 50°C for 10 minutes. Peptides used are listed in Table 1. One of at least three representative experiments is shown.

Next, we optimized the temperature and time for peptide exchange for each allele. For HLA–A*03:01, we tested three template peptides that were unstable at 50°C and could be rescued in the presence of a high-affinity peptide. Thermal stability was assessed by incubating the HLA–A*03:01 complex with a high-affinity peptide (RGAVRATRR) at various temperatures for 1 hour (Figure 2 & Supplemental Figure 2). We selected the template peptide RGAVRATKR, which had the lowest optimal exchange temperature (23°C) for 1 hour. This condition resulted in a nearly complete unfolding of the complex in the absence of a high-affinity peptide and full rescue of the HLA–A*03:01 complex in the presence of a high-affinity peptide (Figure 2). Similar results were obtained when the complexes were incubated at 30°C. A similar approach was used for HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02, where the optimal peptide exchange conditions entailed an 1-hour incubation at 30°C (Figure 2 & Supplemental Figure 2). The peptide exchange efficiency for these HLA alleles was around 80% or higher (data not shown), similar to what was previously shown for HLA-A*02:01 and H2Kb[13].

Optimal exchange conditions for template-peptide MHC-I monomers.

Thermal stability of the template peptide=MHC-I complex to determine optimal exchange temperature with one high-affinity peptide for all alleles tested. Primary data of temperature-based peptide exchange analyzed by gel filtration chromatography at indicated temperatures. Peptides used are depicted in Table 1. One of at least three representative experiments is shown.

Temperature exchange tetramer binding to clonal T cell lines.

Clonal CD8+ T cell lines stained with PE-conjugated temperature exchange multimers that either contained a relevant (pink) or irrelevant peptide (grey). The specificities of the T cell lines are depicted above each figure, as well as HLA-type and first three amino acids of the peptide. Y-axis shows the individually scaled counts and x-axis shows the mean fluorescent intensity of PE.

Temperature-exchanged MHC-I multimers for detection of antigen-specific CD8+ T cells

Previously, we demonstrated that HLA-A*02:01 MHC-I multimers generated via temperature-based peptide exchange performed comparable to those produced using conventional methods[13, 16]. We also established that peptide exchange can be directly applied to multimers themselves, resulting in a more user-friendly protocol with reduced handling time[13, 16]. To evaluate the effectiveness of the newly developed temperature-exchanged MHC-I multimers for detection of antigen-specific T cell responses, we performed temperature exchange reactions using a panel of high-affinity peptides specific for HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02. These temperature-exchanged MHC-I multimers efficiently labelled clonal T cell lines (Figure 3), in a concentration dependent manner (Supplemental Figure 3). This evaluation included two HLA-A*03:01-restricted T cell lines, one HLA-A*11:01-restricted T cell line, two HLA-B*07:02-restricted T cell lines, and one HLA-C*07:02-restricted T cell lines. A complete list of peptides used for temperature-based peptide exchange is provided in Supplemental Table 2. We then compared the staining of T cells by temperature-exchanged MHC-I multimers to those stained with classical folded MHC-I multimers. In the latter case, each MHC-I multimer has to be produced individually, while temperature exchange allows for a batch that can be taken from the freezer and warmed up in the presence of intended peptides before flow analyses. This procedure takes only 1 hr. For this comparison, we generated temperature-exchanged MHC-I multimers for each of the four newly developed alleles and compared their staining performance to that of the classical folded MHC-I multimers at isomass for the same peptide (Figure 4). Additionally, staining intensity was evaluated across multiple dilutions using isomass MHC-I multimer concentrations (Supplemental Figure 4). Overall, the temperature-exchanged MHC-I multimers showed distinct binding to clonal T cells in an antigen-specific manner and exhibited only a minor decrease in performance compared to the classical MHC-I multimers, particularly for the HLA-A11:01 and HLA-B07:02 complexes. Temperature-exchanged multimers for HLA-A*03:01 and HLA-C*07:02 showed similar or slightly lower staining intensities compared to their classical counterparts. Higher concentrations of both multimer types can be used to further enhance staining intensity without any technical complications.

Comparison of temperature exchange multimers to conventional multimers on clonal T cell lines.

Clonal CD8+ T cell lines stained with 13.3 ng/μl of PE-conjugated temperature exchange tetramers (pink) or with conventional tetramers (blue). The specificities of the T cell lines are depicted with the HLA-type and first three amino acids of the peptide for each line. Y-axis shows the individually scaled counts and x-axis shows mean fluorescent intensity of PE.

Detection of virus-specific CD8+ T cells in peripheral blood

To evaluate the practical utility of our additional HLA alleles for temperature-based peptide exchange MHC-I multimers using human peripheral blood samples, we compared the performance of temperature-based peptide exchanged multimers to that of classical multimers within an immune monitoring framework. This is particularly relevant in the context of monitoring virus-specific T cells in immunocompromised patients, such as those undergoing T cell-depleted allogeneic stem cell transplantation and adoptive cell therapy. Herpesviruses, including human cytomegalovirus (HCMV) - and Epstein-Barr virus (EBV), can cause severe infections in these patients. Restoring T cells is essential to prevent these infections and to minimize complications and mortality. We analyzed peptides derived from HCMV and EBV for the HLA types HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02 in peripheral blood samples from healthy donors. Additionally, we included the previously developed temperature-based peptide exchange multimers for HLA-A*02:01 in this analysis[13]. We exchanged PE- and APC-labeled multimers for a selection of these peptides to measure T cell frequencies in peripheral blood mononuclear cells (PBMCs) and compared them to traditional folded multimers. The antigen-specific T cells identified with temperature-exchanged multimers stained a distinct cell population (Figure 5). Importantly, we did not observe any MHC-I multimer positive T cells in two irrelevant donors (Supplemental Figure 5). The percentages observed are comparable to those observed with conventional multimers (Figure 5). The peptides are listed in Supplemental Table 2 and the gating strategy of the immune monitoring framework is shown in Supplemental Figure 6. These result illustrates the efficiency and versatility of our technology in rapidly generating diverse MHC-I multimers from a ready-to-use stock, enabling the detection of antigen-specific T cells, even at the low frequencies commonly observed in primary immune monitoring samples.

Comparison of temperature exchange tetramers to conventional tetramers on PBMCs.

PBMCs of healthy individuals were incubated with peptide exchange (pink) or conventional (blue) tetramers. Donor PBMCs were CMV- or EBV-positive based on serology and were incubated with the relevant tetramer based on donor HLA-type. Tetramers were conjugated to APC (y-axis) as well as PE (x-axis). Gate shows the tetramer-binding CD8+ T cells and the percentage of total CD8+ T cells is depicted in the top right corner.

Discussion

The use of multimeric versions of peptide MHC-I complexes has become an important technology to detect and quantify antigen-specific CD8+ T cells[22-25]. To identify antigen-specific T cells using flow cytometry, fluorescently labeled MHC multimers are widely employed. There is an increasing interest in the detection of CD8+ T cells immune responses since they are important for the protection against pathogens or cancer and can be used to define antigens for vaccination purposes. Two challenges in advancing this field are the need for high-throughput approaches and the lack of ready-to-use reagents. Classical peptide-MHC-I complexes are each produced in a separate production run, required for each unique peptide-HLA complex. The process is elaborate, time consuming and demands highly skilled people. This restricts the generation of a larger library of peptide-HLA combinations and thereby results in incomplete detection of the full CD8+ T cell response against an antigen. Previously, it has been shown that peptide-exchange technologies, such as UV-based peptide exchange, are able to upscale towards higher throughput approaches. However, quantifying the efficiency of UV-induced peptide destruction and subsequent peptide exchange remains challenging. Temperature-based peptide exchange is controlled, simple, fast, and quantitative if the correct conditions are identified, as for the HLA alleles in this manuscript. The time to prepare the materials for flow analyses is around 1 hour and consists of taking the exchange MHC-I multimer from the fridge and warming it up in the presence of peptides of interest. This can be immediately followed by staining the CD8+ T cells for flow cytometric analysis.

Additionally, MHC-I molecules can interact with other surface proteins in both a peptide-dependent and peptide-independent manner. For example, they bind to the inhibitory receptor Leukocyte immunoglobulin-like receptor subfamily B member 1 (LILRB1), independently of the sequence of the presented peptide, while interactions with receptors such as Killer Immunoglobulin-like Receptors and CD55 are peptide-dependent[26-29]. Consequently, a positive flow cytometry signal does not necessarily indicate interaction with the TCR of CD8+ T cells. It is therefore essential to exercise caution when interpreting MHC-I multimer results to ensure accurate conclusions. These diverse MHC-I based interactions may provide further utility for studying various immune cell subsets beyond T cells, including natural killer (NK) cells and other innate immune cells from a peptide MHC-I specific perspective.

We have set out to determine whether temperature-based peptide-exchange technology can be developed into a broadly applicable platform for the screening of human CD8+ T cells. For this purpose, we designed conditional peptide ligands for four different human MHC-I alleles across the three human HLA loci: HLA-A*03:01, HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02. Refolding reactions with the temperature-based peptide are efficient and result in stable peptide MHC-I complexes at 4°C, which are unstable with increased temperature. The presence of an high-affinity peptide during temperature-based peptide exchange ensures the replacement of the temperature conditional ligand in the MHC-I peptide-binding groove. This leads to the generation of the peptide-MHC-I complex with a desired antigen specificity. Importantly, we show that the temperature-based peptide exchange can be performed using the multimerized and fluorescently labeled MHC-I complexes. This allows the development of ready-to-use MHC-I multimers, with high-throughput abilities demonstrating the ease of use of temperature-based peptide exchange technology.

The temperature-based peptide exchange method for generating MHC-I multimers, has several limitations. First, effective peptide exchange necessitates the design of HLA-specific ligands, adding complexity compared to approaches that utilize empty-loadable MHC-I multimers. This reliance on specific ligands could also affect the specificity and functionality of T cell responses, particularly when low-affinity peptides are involved. Additionally, although the method allows for rapid ready-to-use multimer generation, the inherent stability of some MHC-I complexes could limit the types of peptides that can be successfully exchanged. Furthermore, while this system is designed to be accessible to a wide range of laboratories, the initial need for appropriate peptide selection and optimization might still pose a challenge for some users. Lastly, despite the benefits of temperature-based peptide exchange technology, there is still a risk that some combinations may not provide the desired specificity and functionality. This highlights the need for ongoing refinement and validation of the method across different MHC-I alleles. Additionally, using complementary peptide exchange methods may enhance overall effectiveness as has been done for UV-exchange multimers and empty disulfide-stabilized multimers[30].

Current peptide exchange technologies predominantly target HLA alleles common in individuals of Western European descent. However, we have started expanding this work to include HLA-A*11:01, which is most prevalent in East and Southeast Asian populations [31, 32]. Currently there are temperature-based peptide-exchange multimers for six HLA alleles, namely HLA-A*02:01, HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, HLA-C*07:02 and HLA-E[14, 15]. With this set of MHC-I reagents, a deep analysis could be possible for people with an European genetic background.

A variety of MHC-I reagents are currently available for use in CD8+ T cell diagnostics, including classical folded MHC-I multimers, empty stabilized multimers and UV-exchange multimers. These MHC-I reagents are being used to determine virus specific T cell diagnostics, personalized neoantigen-specific T cell diagnostics and more. We propose that the ready-to-use temperature-based peptide exchange MHC-I technology could further advance the field, as these reagents simplify the process by eliminating additional steps typically required for MHC-I multimers production.

Material and methods

PBMCs isolation

PBMCs were isolated from healthy donors and obtained via Sanquin or biobanked by the Department of Hematology (LUMC) after informed consent, in accordance with the Declaration of Helsinki. PBMCs were isolated from buffy coats or whole blood using Ficoll-Isopaque and used directly or stored using cryo-preservation. Biobanked PBMCs were thawed in Iscove Modified Dulbecco Medium (IMDM; Lonza, Basel, Switzerland) supplemented with 10% heat-inactivated fetal bovine serum (FBS; Sigma-Aldrich, Saint Louis, MO, USA), 2.7 mM L-glutamine (Lonza), 100 U/mL penicillin (Lonza), and 100 µg/mL streptomycin (Lonza; 1% p/s), and subsequently treated with 1.33 mg/mL DNAse to minimize cell clumping. PBMCs were either used for isolation of clonal T cell lines or for flow cytometry analysis.

Clonal T cell lines

Clonal T cell lines were isolated and expanded as previously described[33, 34]. In brief, PBMCs were incubated with peptide-HLA multimers and antibodies targeting CD14, CD8 and CD4 to isolate CD14-CD8+CD4-multimer-binding cells using fluorescence-activated cell sorting (FACS). In some cases, cells were pre-selected prior to FACS using magnetic-activated cell sorting (MACS) based on peptide-HLA multimer-binding. Obtained cells were further expanded either as bulk[33] or single cell per well[34], in the presence of irradiated feeders with or without irradiated Epstein-Barr virus lymphoblastoid cell lines (EBV-LCLs), phytohemagglutinin (PHA) or specific peptides. See Supplemental Table 1 for T cell line details.

Flow cytometry

1-2×106 PBMCs or 0.05-0.1×106 T cells were seeded in a round bottom 96-wells plate. Cells were stained with the viability dye Zombie-Red (Biolegend, San Diego, CA, USA) for 25 min at room temperature after which the cells were washed in PBS containing 0.8 mg/mL albumin (FACS buffer). Cells were stained with 20 μl multimer staining mix, incubated for 30 minutes at 4°C. Conventional multimers were generated in-house, as previously described[33]. After washing, cells were stained with 20 μl of an antibody cocktail containing: αCD8-pacific blue (cat#558207, BD Biosciences), αCD4-BV510 (cat#562970, BD Biosciences), αCD14-FITC (cat#555397, BD Biosciences), αCD19-FITC (cat#555412, BD Biosciences) and αCD16-FITC (cat#335035, BD Biosciences). After incubation for 20 minutes at 4°C, cells were washed and measured on a 3-laser Aurora (Cytek Biosciences) or LSR II (BD Biosciences) flow cytometer. Flow cytometry data analysis was performed using OMIQ.ai. See Supplemental Table 2 for patient details regarding CMV- and EBV status and HLA typing.

Peptide synthesis and purification

Peptides were synthesized at the LUMC Peptide Facility by standard solid-phase peptide synthesis using Syro I and Syro II synthesizers. Peptides were purified using reversed-phase HPLC (Waters) using a mobile phase of water/acetonitrile gradient containing 0,1% TFA on a C18 column (X-bridge, Waters). Purity of the peptides was confirmed using LC-MS (Micromass LCT Premier; Waters).

Generation of temperature sensitive monomers and multimerization

Temperature-based sensitive monomers were produced as previously described[16]. Briefly, recombinant HLA-A/B/C heavy chain and B2M light chain were produced as inclusion bodies in bacterial strain BL21 (DE3). These purified inclusion bodies containing 2.5 mg of heavy chain were solubilized in 8 M urea and folded for 5 days at 10°C with 1.2 mg prefolded B2M and 3 mg temperature-sensitive peptide (A3: RGAVRATKR; A11: RFKMFPEVR; B07: RKELVRPAL; C07: IKYNYPAMA) in 50 ml folding buffer (400 mM L-Arginine, 0,5 mM oxidized glutathione, 5 mM reduced glutathione, 2 mM EDTA, 100 mM Tris.HCl pH 8, glycerol 5% and half a tablet of protease inhibitor cocktail (complete, Roche)). Temperature-based sensitive monomers were concentrated on a 30 kDa filter (Amicon Ultra-15) and biotinylated with BirA ligase overnight at 4°C. Biotinylated Temperature-based sensitive monomers were purified with size exclusion chromatography (Superdex 75 10/300 GL) on a NGC system (Bio-Rad). Properly folded complexes were concentrated on a 30 kDa filter (Amicon Ultra-15) and the concentration was measured on a NanoDrop (Thermo Fisher Scientific) and stored in 300 mM NaCl and 20 mM Tris-HCl, pH 8.0 with 15% glycerol at -80°C. Next, the monomers were multimerized with PE-labeled streptavidin and/or APC-Labeled streptavidin (SA-PE, SA-APC; Thermo Fisher Scientific, Invitrogen) to a final concentration of 0,625 µM. For peptide exchange, temperature-based monomers were incubated at a final concentration of 0,5 µM and 50 µM of selected peptide. This was diluted 1:20 with PBS to be added to CD8+ T cells.

HPLC analysis of temperature-mediated peptide exchange

To initiate temperature-based peptide exchange, a 0.5 µM peptide–MHC-I complex was mixed with 50 µM exchange peptide in 110 µl PBS and incubated under defined conditions. After centrifugation at 14,000 g for 1 min, the supernatant was analyzed using size-exclusion chromatography on a Shimadzu Prominence HPLC system with a 300 × 7.8 mm BioSep SEC–s3000 column and PBS as mobile phase. Data were analysed using Shimadzu LabSolutions software.

Glossary

MHC Class I; Peptide Exchange; HLA Alleles; CD8+ T Cells; Temperature-sensitive Multimers; Antigen-specific Immunity; Fluorescent Labeling; High-throughput Technology

eLife Assessment

This study presents an alternative to conventional and UV-based tetramers, which are easy to use and reliable for the identification of antigen-specific CD8 T cells. The authors demonstrate that tetramers for HLA alleles A0301, A1101, B0702, and C0702 can be subjected to specific temperatures that facilitate peptide exchange, whilst maintaining structural integrity. Whilst the strength of the evidence is currently incomplete, further development and validation of this approach is likely to provide a useful alternative to generating reagents for examining T cell specificities.

Reviewer #1 (Public review):

Summary:

A fundamental technique for the identification of peptide-specific CD8 T cells is the use of fluorophore-conjugated and peptide loaded MHC tetramers. Classically, refolding of specific peptides with MHC monomers can be labour intensive, and not optimal for screening large numbers of different peptides. Hence, UV-exchanged tetramers have been developed to upscale this, however, still has some associated challenges such as UV-mediated damage to peptide complexes. Here, Pothast, C.R. et al demonstrate the efficacy of using temperature exchanged tetramers for the prevalent alleles HLA-A*03:01, A*11:01, B*07:02, and C*07:02. Building upon their previous work with HLA-A*02:01, H-2Kb, and HLA-E. They first demonstrate the complex stability of tetramers with different affinity peptides at high temperature, showing complex destabilisation can be rescued with higher affinity peptides. This is followed by an optimisation of peptide exchange temperatures, tailored for each allele. The authors then demonstrate successful binding to clonal T cell lines, and then a step further with viral peptides against PBMCs from individuals with confirmed infection history. For the latter they compare to conventional tetramers and demonstrate comparable signal.

Due to the prevalence of these 4 alleles, the ease-of-handling, and short time requirements, these tetramers are likely to show high utility.

Strengths:

The manuscript is well-written and the results are solid, although more detail may add clarity to some of the results, in particular Figures 1 and 2. Other than the points reported below, the study uses accurate controls to demonstrate the specificity of the tetramers, and the data are convincing.

Overall, the interpretation of the results is accurate, and the discussion is thorough. Additional comments may be included to cover potential tetramer batch variability and differences in the stability of different alleles. Specifically, whether certain alleles require higher-affinity peptides to be stable, compared to others.

Weaknesses:

The authors demonstrate the equivalence of temperature-exchanged tetramers to conventional ones, however, as they are an advancement on UV-exchange, it would be useful to show data on how their stability, exchange efficacy, and binding to T cell lines compare to UV-based tetramers. It would be supportive to show that temperature does not impact fluorophore intensity as well.

Reviewer #2 (Public review):

Summary:

The majority of CD8+ T cell responses rely on the proper presentation of antigens through stable MHC-I (but not requiring a stable immunological synapse). This work highlights a new approach to build an array of stable peptide MHC-I using temperature exchange, which can be used to identify antigen-specific CD8+ T cells.

Strengths:

In this work, the authors have proposed an alternative method to reload the peptide MHC-I molecule. Their temperature-exchange approach is distinct from current reloadable peptide MHC technologies involving photolabile peptide, empty MHC-I (Nat Commun 11, 1314 (2020). https://doi.org/10.1038/s41467-020-14862-4), tapasin/TAPBPR chaperone-assisted (eLife 7:e40126.), enzyme exchangeable (WO2020226570) and small alcohol (Curr Res Immunol. 2022 Aug 18;3:167-174. doi: 10.1016/j.crimmu.2022.08.002) approaches.

Weaknesses:

However, the proposed temperature-exchange approach does not substantially improve the quality of antigen-specific T cells that can be identified using the photolabile peptide MHC-I molecules.

The time saved using the temperature-exchange protocol may not be a pull factor as the photolabile peptide MHC-I approach is not unreasonably laborious.

Reviewer #3 (Public review):

Summary:

The study by Pothast and colleagues outlines an extension of their previously described temperature-based MHC-I peptide exchange method on 4 common HLA alleles, to enable the generation of peptide/MCH-I tetramers for characterization of antigen-specific T cells by flow cytometry.

Strengths:

This work outlines a protocol for generating MHC-I tetramers on 4 common HLA allotypes, which can then be applied to monitor T cell responses by flow cytometry studies. The work provides conditional ligands for exchange on each HLA and demonstrates proof of concept studies using clonotypic T cells and CD8+ PBMCs.

The results support that the temperature-exchanged tetramers can perform similarly to conventional tetramers in some settings.

Weaknesses:

Given that there are several proposed methodologies addressing the same task (including UV-mediated, disulfide-bond based stabilization of empty MHC-I conformers, and chaperone-based methods), the relevance of the proposed temperature-mediated technology is questionable.

More specifically, important limitations of the study include:

(1) A lack of quantification of exchanged molecules relative to molecules that retain the original placeholder peptides, or completely empty molecules present in the same sample.

(2) A lack of validation that peptide exchange has occurred in the absence of a reporter T cell line appears to be a significant limitation of the methodology for antigen / T cell discovery.

(3) The sub-optimal exchange efficiency relative to conventional prepared pMHC-I molecules, shown in Figure 4, is a significant limitation of the approach.

(4) There are no data to support that exchange proceeds through the generation of empty molecules during the temperature cycle, or by peptide binding on empty molecules that are already present in the sample. Understanding the mechanism of exchange is important for the necessary improvements to the methodology.

(5) It is possible that the temperature cycle causes protein aggregation or other irreversible changes to the sample - this should be explicitly quantified and addressed in the paper, since misfolded MHC-I molecules can lead to high levels of background staining.

(6) These potential limitations should limit detection of low-affinity/low-avidity interactions between TCRs and their cognate pMHC antigens - this should be addressed explicitly in a model antigen setting.

(7) The approach appears to be limited to the HLAs showing high thermal stability, which have been explored in this study. However, a large fraction of HLAs show sub-optimal thermal stabilities. It seems that explicit validation of peptide exchange would be required for any new HLA allele introduced into this process.

(8) Whether the approach can be used to load suboptimal peptides with lower thermal stabilities that are emerging immunotherapy targets is not addressed in the present study.

Because of these limitations, the present manuscript does not conclusively support the claim that temperature-based exchange can be used as a robust methodology to generate pMHC-I tetramers with desired peptide specificities.

As a result, the scope of applications using these suboptimal exchanged pHLA tetramers is limited, and should be addressed with further improvements of the methodology, including better characterization of exchange efficiency, demonstration of functionality across a broader range of HLA allotypes with varying thermal stability profiles, and validation with clinically relevant low-affinity peptides that would strengthen the potential utility of this approach in immunotherapy development and basic T cell biology research.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

A fundamental technique for the identification of peptide-specific CD8 T cells is the use of fluorophore-conjugated and peptide loaded MHC tetramers. Classically, refolding of specific peptides with MHC monomers can be labour intensive, and not optimal for screening large numbers of different peptides. Hence, UV-exchanged tetramers have been developed to upscale this, however, still has some associated challenges such as UV-mediated damage to peptide complexes. Here, Pothast, C.R. et al demonstrate the efficacy of using temperature exchanged tetramers for the prevalent alleles HLA-A*03:01, A*11:01, B*07:02, and C*07:02. Building upon their previous work with HLA-A*02:01, H-2Kb, and HLA-E. They first demonstrate the complex stability of tetramers with different affinity peptides at high temperature, showing complex destabilisation can be rescued with higher affinity peptides. This is followed by an optimisation of peptide exchange temperatures, tailored for each allele. The authors then demonstrate successful binding to clonal T cell lines, and then a step further with viral peptides against PBMCs from individuals with confirmed infection history. For the latter they compare to conventional tetramers and demonstrate comparable signal.

Due to the prevalence of these 4 alleles, the ease-of-handling, and short time requirements, these tetramers are likely to show high utility.

Strengths:

The manuscript is well-written and the results are solid, although more detail may add clarity to some of the results, in particular Figures 1 and 2. Other than the points reported below, the study uses accurate controls to demonstrate the specificity of the tetramers, and the data are convincing.

Overall, the interpretation of the results is accurate, and the discussion is thorough. Additional comments may be included to cover potential tetramer batch variability and differences in the stability of different alleles. Specifically, whether certain alleles require higher-affinity peptides to be stable, compared to others.

Weaknesses:

The authors demonstrate the equivalence of temperature-exchanged tetramers to conventional ones, however, as they are an advancement on UV-exchange, it would be useful to show data on how their stability, exchange efficacy, and binding to T cell lines compare to UV-based tetramers. It would be supportive to show that temperature does not impact fluorophore intensity as well.

Reviewer #2 (Public review):

Summary:

The majority of CD8+ T cell responses rely on the proper presentation of antigens through stable MHC-I (but not requiring a stable immunological synapse). This work highlights a new approach to build an array of stable peptide MHC-I using temperature exchange, which can be used to identify antigen-specific CD8+ T cells.

Strengths:

In this work, the authors have proposed an alternative method to reload the peptide MHC-I molecule. Their temperature-exchange approach is distinct from current reloadable peptide MHC technologies involving photolabile peptide, empty MHC-I (Nat Commun 11, 1314 (2020). https://doi.org/10.1038/s41467-020-14862-4), tapasin/TAPBPR chaperone-assisted (eLife 7:e40126.), enzyme exchangeable (WO2020226570) and small alcohol (Curr Res Immunol. 2022 Aug 18;3:167-174. doi: 10.1016/j.crimmu.2022.08.002) approaches.

Weaknesses:

However, the proposed temperature-exchange approach does not substantially improve the quality of antigen-specific T cells that can be identified using the photolabile peptide MHC-I molecules.

The time saved using the temperature-exchange protocol may not be a pull factor as the photolabile peptide MHC-I approach is not unreasonably laborious.

Reviewer #3 (Public review):

Summary:

The study by Pothast and colleagues outlines an extension of their previously described temperature-based MHC-I peptide exchange method on 4 common HLA alleles, to enable the generation of peptide/MCH-I tetramers for characterization of antigen-specific T cells by flow cytometry.

Strengths:

This work outlines a protocol for generating MHC-I tetramers on 4 common HLA allotypes, which can then be applied to monitor T cell responses by flow cytometry studies. The work provides conditional ligands for exchange on each HLA and demonstrates proof of concept studies using clonotypic T cells and CD8+ PBMCs.

The results support that the temperature-exchanged tetramers can perform similarly to conventional tetramers in some settings.

Weaknesses:

Given that there are several proposed methodologies addressing the same task (including UV-mediated, disulfide-bond based stabilization of empty MHC-I conformers, and chaperone-based methods), the relevance of the proposed temperature-mediated technology is questionable.

More specifically, important limitations of the study include:

(1) A lack of quantification of exchanged molecules relative to molecules that retain the original placeholder peptides, or completely empty molecules present in the same sample.

(2) A lack of validation that peptide exchange has occurred in the absence of a reporter T cell line appears to be a significant limitation of the methodology for antigen / T cell discovery.

(3) The sub-optimal exchange efficiency relative to conventional prepared pMHC-I molecules, shown in Figure 4, is a significant limitation of the approach.

(4) There are no data to support that exchange proceeds through the generation of empty molecules during the temperature cycle, or by peptide binding on empty molecules that are already present in the sample. Understanding the mechanism of exchange is important for the necessary improvements to the methodology.

(5) It is possible that the temperature cycle causes protein aggregation or other irreversible changes to the sample - this should be explicitly quantified and addressed in the paper, since misfolded MHC-I molecules can lead to high levels of background staining.

(6) These potential limitations should limit detection of low-affinity/low-avidity interactions between TCRs and their cognate pMHC antigens - this should be addressed explicitly in a model antigen setting.

(7) The approach appears to be limited to the HLAs showing high thermal stability, which have been explored in this study. However, a large fraction of HLAs show sub-optimal thermal stabilities. It seems that explicit validation of peptide exchange would be required for any new HLA allele introduced into this process.

(8) Whether the approach can be used to load suboptimal peptides with lower thermal stabilities that are emerging immunotherapy targets is not addressed in the present study.

Because of these limitations, the present manuscript does not conclusively support the claim that temperature-based exchange can be used as a robust methodology to generate pMHC-I tetramers with desired peptide specificities.

As a result, the scope of applications using these suboptimal exchanged pHLA tetramers is limited, and should be addressed with further improvements of the methodology, including better characterization of exchange efficiency, demonstration of functionality across a broader range of HLA allotypes with varying thermal stability profiles, and validation with clinically relevant low-affinity peptides that would strengthen the potential utility of this approach in immunotherapy development and basic T cell biology research.

We are submittng a revised version that addresses the major points raised. We believe that the manuscript in its present form clearly communicates the development of a robust and versatile temperature-based peptide exchange platform applicable across multiple HLA class I alleles, with relevance for both fundamental and clinical immunology. While we appreciate the constructive feedback provided, we feel that the core findings and conclusions are already well supported and suitable for publication.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The main recommendation here is to include a stability and staining comparison to UV-exchanged tetramers, as this will directly address why temperature-exchanged tetramers are superior. It could also be discussed that research groups may have better access to temperature-regulating equipment than UV. Other minor corrections may include:

We thank the reviewer for this suggestion. We have previously demonstrated that HLA‑A*02:01 and H‑2Kb remain stable as frozen monomers, and similarly showed the stability of thermal peptide exchange for HLA‑E, both at the monomer stage and after multimerization into thermal‑exchange HLA‑E tetramers. In addition, we have directly compared these reagents with conventional tetramers (see Figure 5). Conventional monomers and tetramers are routinely stored at −80 °C for several months without loss of staining performance, and we applied the same storage conditions to thermal‑exchange HLA‑E tetramers. Consistent with conventional tetramers, thermal‑exchange tetramers maintained their ability to stain TCR‑expressing cells for up to six months, with optimal staining preserved at both −20 °C and −80 °C. These findings have been published previously and are discussed in the current manuscript.

We would like to emphasize that we do not conclude that thermal peptide exchange is superior to existing technologies. Our intention is to demonstrate that this approach is feasible and to present its advantages and limitations, which are discussed in detail in the manuscript.

In the discussion we have included a sentence about the accessibility of heat exchange in relation to the equipment needed for UV exchange.

“Moreover, the broader availability and simplicity of equipment needed for thermal exchange compared to UV sources enables wider adoption of this approach.”

(1) Greater explanation of the different coloured curves in Figures 1 and 2, and whether all curves represent samples that are incubated at 50 degrees Centigrade.

We have now clarified in the figure legends that the different coloured curves represent the distinct peptide–HLA complexes analysed, and we explicitly state which curves correspond to samples incubated at 50 °C.

(2) Addition of X-axis title for Figures 1 and 2.

An X‑axis title was already present in Figures 1 and 2, but it was minimal; we have now revised it to make the labeling clearer and more informative.

(3) Clarify whether the background signal (negative) increases with the addition of more tetramer in Supplementary Figure 3 (non-specific binding).

The background (negative) signal did not appreciably increase with the addition of higher tetramer concentrations, indicating minimal non-specific binding under these conditions. This is consistent with observations in the MHC class I multimer field, where non-specific binding is generally limited under optimized staining conditions, and MHC class I multimers can be reliably used to detect antigen-specific T cells.

(4) Address stability of different alleles, or whether they are similar.

Peptide stability varies for each peptide–HLA combination rather than being strictly allele‑specific. We have therefore clarified this point in the manuscript. In addition, as requested by the reviewer, we have included a dedicated section describing the stability of the thermal‑exchange template peptides.

“The peptides that formed stable complexes at 4°C but unstable complexes at 20–32°C and 50°C showed affinities ranging from 4 to 11 µM. Such low affinities are uncommon for endogenous peptides and are unlikely to have physiological relevance, given that the complexes are highly unstable at 37°C.”.

(5) Measure fluorescence signal of tetramer before and after temperature exchange, compared to UV, to show fluorophore stability.

Fluorescent streptavidin forms a highly stable complex. During UV-mediated peptide exchange, reagents are exposed to heat generated by the UV lamp, which can partially denature MHC-I proteins. For this reason, it is necessary to centrifuge the complexes after UV exchange to remove denatured protein aggregates. Importantly, the fluorescence of streptavidin itself is not affected by this process. In contrast, thermal exchange is performed at substantially lower temperatures than UV-mediated exchange and does not compromise MHC-I protein integrity. Therefore, in our view, it does not pose a significant risk to fluorophore stability. In addition, UV may destroy fluorescence by simple bleaching, which cannot be an issue with temperature-dependent peptide exchange.

(6) Use HPLC to show stability of tetramer following temperature exchange versus UV.

HPLC analysis of tetramers is not possible and this is also not performed by others. The reason is that the tetramers are very heterogeneous (clustering of streptavidin together and the amount of MHCI monomers attached to streptavidin). As discussed above we do not claim to be better or worse compared to UV-exchange. This technology has a some additional benefits compared to UV-exchange, such as the tetramerization can be done before the peptide exchange. Additional it does not need a calibrated UV lamp but just an heatblock is needed for the peptide exchange making the exchange less complicated and in our opinion easier to implement. The UV exposure may oxidize amino acids, which is unlike to occur under temperature dependent conditions.

Reviewer #2 (Recommendations for the authors):

Minor in figure:

The authors should explain what "inputs" are in the figure legend.

We have now clarified in the figure legends that the different coloured curves represent the distinct peptide–HLA complexes analysed, and we explicitly state the nature of the “inputs”.

Minor in text:

The authors should also state two important limitations when designing temperature-sensitive peptide MHC-I, which are peptide binding affinity and subsequent T cell recognition. Poorly designed peptides can result in unexchanged peptide MHC-I complexes that may be recognized by the CD8 T cells, leading to false positives.

We have adjusted this in the text (discussion) although we did not make poorly designed peptides. The peptide presented in this manuscript are only stable at 4°C and unstable from 20°C and higher. As the T cell staining are performed at room temperature, there will be no peptide MHC-I complex of the original thermal-exchange peptide present as this is disintegrated. Only the stable, higher-affinity peptide-exchanged MHC-I is in complex. This does pose a problem for peptide exchange with low-affinity peptides, which is mentioned in the current version. However, this does not differ from the other loading technologies as in each case the mother peptide has to be exchanged for a proper peptide and a low-affinity peptide will be lost in time in each other case as well

The authors should include other recent reloadable peptide MHC technologies, e.g. chaperone-assisted, enzyme-exchangeble, and small alcohol treatment.

We apologize for not including these technologies and in the current manuscript they are included in the introduction. (see text below)

“Several technologies have since been developed to allow the parallel production of multiple MHC-I multimers, facilitating high-throughput analysis. For instance, dipeptides have been used as catalysts for peptide exchange[7-10][FS1] , UV-sensitive conditional MHC ligands have been a widely adopted approach [11, 12], in which a photocleavable peptide is cleaved upon UV exposure. Additional peptide exchange technologies include: chaperone-mediated peptide exchange[13, 14] enzymatic cleavage of low-affinity peptides that are covalently linked [15] and small alcohol peptide exchange [16].”

Major in data:

The authors should show if any T cell was detected using the original unexchanged peptide MHC-I in cytometry, as the interpreted use of an irrelevant peptide also depends on the existence of antigen-specific T cells in circulation.

The unexchanged peptide–MHC‑I complex cannot be used for T‑cell staining under our standard conditions because the template peptide dissociates at room temperature, leading to disintegration of the complex. Since our T‑cell staining protocol is performed at room temperature, unexchanged complexes would not remain intact long enough to allow reliable detection. Performing this experiment would require modifying the standard protocol and carrying out all staining steps strictly at 4°C. We choose not to change the staining protocol to perform this experiment.

In addition, the thermal‑exchanged peptide–MHC‑I complexes do not produce elevated background staining. As shown in Figure 5, background staining would appear as single‑positive events (PE or APC only), which we do not observe. This indicates that the thermal exchange approach does not introduce nonspecific tetramer binding.

Reviewer #3 (Recommendations for the authors):

(1) The proportion of exchanged molecules should be quantified relative to those retaining the original placeholder peptides and completely empty molecules present in the sample, using quantitative mass-spectrometry analysis. This would provide critical information about exchange efficiency and help assess the method's reliability.

We thank the reviewer for this point. We have performed this experiment and quantified the efficiencies. This is in the current version in figure 2b.

(2) Validation studies demonstrating peptide exchange in the absence of reporter T cell lines should be included. This would strengthen the methodology's applicability for antigen/T cell discovery applications where such reporter lines may not be available.

We thank the reviewer for raising this important point. Indeed, incomplete or unstable peptide exchange is a general concern inherent to all MHC-I peptide exchange protocols. To address this, we implemented a T cell–independent validation strategy based on LILRB1 binding. The innate immune receptor LILRB1 selectively recognizes properly folded peptide–MHC-I complexes by interacting with β2-microglobulin (β2M) and the α3 domain of the MHC heavy chain in a peptide-independent manner.

Critically, in our system the conditional template peptides are thermally unstable and dissociate during the exchange step. As a result, only successfully exchanged complexes containing a stabilizing peptide remain intact. Consequently, LILRB1 binding serves as a functional readout of successful peptide exchange and pMHC stability.

To assess this, we used K562 cells retrovirally transduced with LILRB1 and stained them with the corresponding fluorophore-conjugated pMHC multimers. Flow cytometric analysis enables clear discrimination between stable (successfully exchanged) and unstable complexes.

This assay is fully independent of TCR-based reporter systems and can therefore be broadly applied in antigen discovery settings where such reagents are not available. Importantly, it complements our HPLC-based quantification of peptide exchange efficiency (Figure 2B) by providing a scalable quality-control readout of structural integrity (Figure 2C).

We have now included this analysis in the Results section and updated the Discussion and Materials and Methods accordingly.

(3) Optimization strategies to improve exchange efficiency should be explored, along with direct comparisons with established methods.

In the revised manuscript, we have included the quantified exchange efficiencies of the thermal peptide exchange reaction, as determined by HPLC analysis on monomeric complexes. In addition, we now present a novel assay that enables routine quality control of each peptide exchange. This analysis is performed using LILRB1+ cells and can be readily applied to any peptide–MHC complex using standard flow cytometry.

Our data demonstrate that the exchange efficiencies obtained with our method are robust, and for the HLA alleles included in this study no further optimization was required. With respect to comparisons with established technologies, we chose to benchmark our approach against conventionally folded tetramers (Fig. 5). As stated above, we do not claim superiority over other peptide-exchange platforms; rather, we aim to present an alternative workflow with distinct practical advantages.

(4) Potential protein aggregation or other irreversible structural changes resulting from temperature cycling should be quantified and addressed. Since misfolded MHC-I molecules can contribute to high background staining, this characterization is essential for assessing the method's reliability. This can be done using DLS studies, which should report on the fraction of misfolded, high MW species present in the sample upon temperature treatment.

We thank the reviewer for raising this important point regarding potential aggregation or irreversible structural changes during temperature cycling. We agree that the presence of misfolded or aggregated MHC-I complexes could contribute to nonspecific staining and affect the reliability of the method.

Although we did not perform dynamic light scattering (DLS), we addressed this concern using complementary biochemical and functional approaches. First, monomeric peptide–MHC complexes were analyzed by size-exclusion HPLC throughout the study. These analyses consistently showed well-defined peaks corresponding to properly folded complexes, without detectable high–molecular weight species that would indicate aggregation (Figures 1 and 2).

Second, and importantly, we assessed the functional consequences of any potential misfolding or aggregation at the multimer level. If aggregated or misfolded MHC-I complexes were present, this would be expected to result in increased nonspecific binding and elevated background staining. However, across all experiments using both clonal T cell lines and primary PBMCs, we consistently observed low background staining and clear antigen-specific populations (Figures 3–5), arguing against the presence of aggregation-prone or sticky multimer species.

Third, we incorporated an additional quality control step based on LILRB1 binding (Figure 2C). Because LILRB1 selectively recognizes properly folded peptide–MHC-I complexes, this assay provides a sensitive readout for structural integrity. Unstable or improperly folded complexes do not bind LILRB1, allowing discrimination between functional and non-functional pMHC complexes.

Taken together, these complementary analyses indicate that temperature-based peptide exchange does not induce detectable aggregation or irreversible structural damage under our experimental conditions, and that the resulting multimer reagents are structurally intact and suitable for downstream applications. We have clarified this in the revised manuscript.

(5) The method's capacity to detect low-affinity/low-avidity TCR-pMHC interactions in a well-characterized model antigen system should be explicitly addressed. This would define the sensitivity limits of the approach compared to established methods.

The reviewer raises a valid and broadly relevant point that applies to all MHC-based technologies. However, we consider a full discussion of this topic to be outside the scope of the current manuscript. We do address this limitation for low-affinity peptides in the Discussion, where we write: “This technology is most likely not optimal for low-affinity peptides, as the template peptide itself must also have low affinity.” However, this problem may be universal since low-affinity peptides will simply release all MHC class I-peptide exchanged complexes during the staining process.

(6) Since the current study focuses on HLAs with high thermal stability, the approach should be tested with HLA alleles exhibiting lower thermal stability profiles. A validation framework could be developed that could be applied when introducing new HLA alleles to this process.

We thank the reviewer for this insightful comment. To our knowledge, there is no established classification of HLA alleles strictly based on “high” versus “low” intrinsic thermal stability; rather, stability is largely determined by the peptide–HLA combination.

Regarding the development of a validation framework for additional HLA alleles, our data indicate that suitable template peptide affinities generally fall within a range of approximately 4–11 µM. We anticipate that this range may vary between alleles, and therefore each HLA allele will require empirical identification of an optimal template peptide to achieve appropriate thermal stability and efficient peptide exchange.

This can be accomplished using the workflow presented in our study: first, assessing complex stability at 50 °C in the absence or presence of a high-affinity peptide (Fig. 1), followed by stepwise temperature reduction to determine optimal exchange conditions and incubation times (Fig. 2). In addition, LILRB1-based quality control can be incorporated to confirm the formation of properly folded and stable peptide–MHC complexes.

(7) It is encouraged that the approach be explored for its ability to effectively load suboptimal peptides with lower thermal stabilities. Demonstrating this capability would significantly enhance the method's clinical relevance.

We agree with the reviewer that evaluating the performance of the approach on suboptimal, low‑affinity peptides would be highly valuable and would further strengthen the clinical relevance of the method. In the current study, however, this falls outside our experimental scope. We are working on this in a follow-up project. We have now acknowledged the limitation of low-affinity peptides in the revised manuscript, noting that low‑affinity peptides are less suitable to this technology because the template peptide itself must also possess low affinity. Future work will be needed to systematically explore template design and exchange conditions optimized for such peptides.

Version 2

Abstract

T cell recognition of specific antigens presented by major histocompatibility complexes class-I (MHC-I) can play an important role during immune responses against pathogens and cancer cells. Detection of T cell immunity is based on assessing the presence of antigen-specific cytotoxic CD8+ T cells using MHC class-I (MHC-I) multimer technology. Previously we have designed conditional peptides for HLA-A*02:01, H-2Kb and HLA-E that form stable peptide-MHC-I-complexes at low temperatures and dissociate when exposed to a defined elevated temperature. The resulting conditional MHC-I complex can easily and without additional handling be exchanged with a peptide of interest, allowing to exchange peptides in a ready-to-use multimer and a high-throughput manner. Here we present data that this peptide-exchange technology is a general applicable, ready-to-use and fast approach to load many different peptides in MHC-I multimers for alleles of the HLA-A, HLA-B and HLA-C loci. We describe the development of conditional peptides for HLA-A*03:01, HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02 that only form stable peptide-MHC-I complexes at low temperatures, allowing peptide exchange at higher defined temperature. We document the ease and flexibility of this technology by monitoring CD8+ T cell responses to virus-specific peptide-MHC complexes in patients.

Introduction

An essential component of cellular immunity is mediated through the T Cell Receptor (TCR) of CD8+ T cell that can bind MHC-I complexes. The MHC-I complex consist of three components: a polymorphic heavy chain, the β2-microglobulin (β2M) protein, and an 8–10-amino-acid peptide ligand[1]. Peptide selection occurs through docking of the peptide C-terminus in MHC-I and occupation of the pockets in MHC-I followed by a conformational change that preferably traps the peptides with highest MHC-binding affinity. As a result, peptides with low MHC-binding affinity are replaced by peptides with a high MHC-binding affinity, a phenomenon used in this work [2]. The peptide ligand is an essential part of the MHC-I complex; without it, the complex cannot remain stable and dissociate[3]. Pathogens, such as viruses, can infect human cells and a primary infection often leads to disease and immunity. Recurrent infections typically do not cause severe disease and are cleared, in part by T cells recognizing viral peptides presented by MHC-I, as recently demonstrated for SARS-CoV-2[4, 5]. The emergence of viral variants can potentially give rise to a partial change in immunogenic viral peptides thereby affecting disease progression. Moreover, immune-compromised patients who lack proper B cell immunity are fully dependent on T cell-mediated immunity. Therefore, monitoring antigen-specific T cells is crucial for tracking the impact of recurrent infections and determining whether therapeutic intervention is required.

The detection and visualization of antigen-specific T cells was first achieved by using peptide-loaded MHC-I multimers with a fluorochrome[6]. To generate such labelled multimers, soluble MHC-I monomers complexed with a peptide are multimerized through biotinylating and fluorochrome-conjugated streptavidin. This process is time-consuming and involves multiple steps for each individual peptide-MHC-I complex. Several technologies have since been developed to allow the parallel production of multiple MHC-I multimers, facilitating high-throughput analysis. For instance, dipeptides have been used as catalysts for peptide exchange[7–10], UV-sensitive conditional MHC ligands have been a widely adopted approach[11, 12], in which a photocleavable peptide is cleaved upon UV exposure. Additional peptide exchange technologies include: chaperone-mediated peptide exchange[13, 14] enzymatic cleavage of low affinity peptide that are covalently linked [15] and small alcohol peptide exchange [16]. This results in an unloaded MHC-I monomer, which can subsequently be loaded with an high-affinity peptide and multimerized[11, 12]. However, UV exposure can cause photodamage to the peptide MHC-I complex, that leads to sample evaporation, and is incompatible with fluorescently labeled multimers, requiring multimerization to be performed after the peptide exchange[17]. Additionally, more recent methods involve empty MHC-I stabilized by the introduction of a disulfide bond, which can be loaded with the desired a peptide of choice. It is currently unknown how these mutations and structural alterations affect the peptide binding repertoire of these disulfide-stabilized MHC-I molecules[8, 9].

One major limitation of current MHC-I multimer-based technologies is the complex, multistep protocols to obtain ready-to-use peptide MHC-I complexes, without inducing mutations in the HLA class I heavy chain or Β2M. This limits the broad clinical usage of MHC-I multimer technology. To address this, we previously developed a temperature-based peptide exchangeable MHC-I multimer technology for human HLA-A*02:01, HLA-E and murine H2Kb, which simplifies and accelerates peptide exchange on MHC-I multimers[18–20]. This technology allows the generation of multimers within hours with ease, simply by incubating selected peptides at a defined temperature with a single standardized batch of fluorescently labeled MHC-I multimers [21]. The MHC-I multimers are multimerized and fluorescently labeled prior to peptide exchange, reducing handling for the end user. In this study, we aimed to determine whether this technology could be developed into a robust and broadly applicable system for analyzing CD8+ T cell immunity. To this end, we developed temperature-based peptide-exchange MHC-I multimers for human MHC gene product HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02, representing all major MHC-I gene families.

Results

Design of temperature-exchangeable template peptide

We set out to extend our toolbox of HLA alleles that allow temperature-based template peptide exchange using multimerized, ready-to-use MHC-I complexes. Our goal was to design peptides that form stable complexes with MHC-I at low temperature and are unstable at higher temperatures, allowing replacement of the template peptide by a desired high-affinity peptide. The peptide off-rate is the primary determinant for the MHC-I stability[22]. For HLA–A*03:01, we designed peptides based on a Hepatitis C virus epitope (HCV43-51; RLGVRATRK) (dissociation constant [Kd] 11.7 nM)[23] and modified the residues to decrease the dissociation constant[24]. HLA–A*03:01 complexes were produced with wild-type or modified peptide, and thermal stability was assessed at 50°C using analytical size exclusion high-performance liquid chromatography (HPLC). With the modified peptide, RGAVRATRR (Kd 7098.44 nM), the complex was unstable at 50°C, resulting in denaturation, as shown by the absence of a MHC-I peak in a size-exclusion HPLC (Figure 1 & Supplemental Figure 1). In contrast, the presence of the high-affinity epitope (RLGVRATRK, 11.7 nM) produced a distinct peak, demonstrating that the HLA-A*03:01 complex could be rescued from unfolding. A similar approach was applied to HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02. For each allele, we designed peptides for each respective allele and modified the residues to decrease the affinity (Table 1)[23, 25, 26]. HLA complexes with the modified peptides were produced, and their thermal stability was evaluated at 50°C using analytical size-exclusion HPLC. For each allele, we identified a thermal template peptide that was unstable at 50°C, which could be rescued by the addition of a high-affinity peptide (Figure 1 & Supplemental Figure 1). We proceeded with peptides that exhibited thermo-instability at 50°C that showed a rescued HLA complex in the presence of the high-affinity peptide. The peptides that formed stable complexes at 4 °C but unstable complexes at 20–32 °C and 50 °C displayed affinities ranging from 4 to 11 µM. Such low affinities are uncommon for endogenous peptides and are unlikely to have physiological relevance, given that the complexes are highly unstable at 37 °C.

Thermal stability of MHC-I monomers with template peptide.

Thermal stability of HLA-A03:01, HLA-A11:01, HLA-B07:02, HLA-C07:01, and HLA-C*07:02 complexed with their template peptides was assessed at 50 °C using analytical size-exclusion HPLC. Peptide–MHC I monomers were incubated with indicated peptides at 50°C for 10 minutes. Peptides used are listed in Table 1. One of at least three representative experiments is shown. The black line represents primary data without thermal exchange, the green line shows thermal exchange without peptide (resulting in complex instability), and the pink line represents thermal exchange in the presence of peptide (yielding a stable complex).

Next, we optimized the temperature and time for peptide exchange for each allele. For HLA–A*03:01, we tested three template peptides that were unstable at 50°C and could be rescued in the presence of a high-affinity peptide. Thermal stability was assessed by incubating the HLA–A*03:01 complex with a high-affinity peptide (RGAVRATRR) at various temperatures for 1 hour (Figure 2 & Supplemental Figure 2). We selected the template peptide RGAVRATKR, which had the lowest optimal exchange temperature (23°C) for 1 hour. This condition resulted in a nearly complete unfolding of the complex in the absence of a high-affinity peptide and full rescue of the HLA–A*03:01 complex in the presence of a high-affinity peptide (Figure 2). Similar results were obtained when the complexes were incubated at 30°C. A similar approach was used for HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02, where the optimal peptide exchange conditions entailed an 1-hour incubation at 30°C (Figure 2 & Supplemental Figure 2). The exchange efficiency was then examined in detail for all 4 of the alleles included in this study. We found that replacing the template peptide with the high-affinity peptide and incubating for approximately 1–1.5 hours at the specified temperature resulted in nearly complete peptide exchange, as shown in Figure 2B. This pattern was consistent across the different HLA alleles tested and aligns with the high exchange efficiencies previously documented for HLA-A*02:01 and H2Kb[18]. For 3 alleles, the measured exchange efficiency exceeded 100%. This effect arises from the intrinsic instability of the input complex at temperatures above 4°C, which partially dissociates during the short preparation and measurement steps. In the condition when a high affinity peptide is present the peptide stabilizes the complex. Thus, the efficiency of thermal exchange may be underestimated due to technical issues. Collectively, these findings establish robust conditions for efficient, thermal peptide exchange across multiple HLA class I alleles and expand the applicability of thermal peptide exchange for multimer production.To verify the quality of the thermally exchanged peptide–MHC (pMHC) complexes, we incorporated an additional control based on LILRB1 binding. Stability of each pMHC complex was assessed using K562 cells retrovirally transduced to overexpress LILRB1 (Fig. 2C). Because LILRB1 recognizes structural elements of β2-microglobulin (β2M) and the α3 domain of the MHC heavy chain, it selectively interacts with properly assembled and stable pMHC complexes. We have previously demonstrated this for both thermally exchanged HLA-E tetramers and UV-mediated peptide exchange reagents [27, 28]. Consistent with this, four conventional PE-conjugated tetramers stained only K562-LILRB1OE cells (Fig. 2D). We then applied the thermal peptide-exchange tetramers, representing four HLA alleles. When a high affinity peptide formed a stable complex with the PE-conjugated tetramer, robust staining of K562-LILRB1OE cells was observed (Fig. 2E). In the absence of peptide, unstable association with the MHC heavy chain led to β2M dissociation from the fluorescent tetramer and no LILRB1-dependent staining was observed. Collectively, these optimizations and validation steps confirm that thermal peptide exchange is a robust and versatile method for generating stable pMHC complexes, enabling reliable downstream detection of antigen-specific T cells.

Optimal exchange conditions for template-peptide MHC-I monomers.

A) Thermal stability of the template peptide-MHC-I complex to determine optimal exchange temperature with one high-affinity peptide for all alleles tested. Primary data of temperature-based peptide exchange analyzed by gel filtration chromatography at indicated temperatures. The black line represents primary data without thermal exchange, the green line shows thermal exchange without peptide (resulting in complex instability), and the pink line represents thermal exchange in the presence of peptide (yielding a stable complex). The thermal exchange temperatures corresponding to each histogram are indicated above the panels and are 23, 30, 30, and 30 °C, respectively. Peptides used are depicted in Table 1. One of at least three representative experiments is shown. B) The exchange efficiency was calculated from the area under the curve measured by HPLC and normalized to binding of the optimal peptide and timing as indicated in each histogram. Mean values ± SD from three independent experiments are shown. C) LILRB1 over-expression (OE) in K562. Grey: negative control, Purple line: K562-WT plus anti-LILRB1-Fitc, Purple filled: K562-LILRB1OE plus anti-LILRB1-Fitc (N=3). D) 4 different conventional tetramer staining on K562-LILRB1OE. Open blackline: negative control. (N-6) E) thermal exchange using a high affinity peptide or no peptide, followed by staining of the K562-LILRB1OE and flow cytometric analysis (N=4).

Temperature-exchanged MHC-I multimers for detection of antigen-specific CD8+ T cells

Previously, we demonstrated that HLA-A*02:01 MHC-I multimers generated via temperature-based peptide exchange performed comparable to those produced using conventional methods[18, 21]. We also established that peptide exchange can be directly applied to multimers themselves, resulting in a more user-friendly protocol with reduced handling time[18, 21]. To evaluate the effectiveness of the newly developed temperature-exchanged MHC-I multimers for detection of antigen-specific T cell responses, we performed temperature exchange reactions using a panel of high-affinity peptides specific for HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02. These temperature-exchanged MHC-I multimers efficiently labelled clonal T cell lines (Figure 3), in a concentration dependent manner (Supplemental Figure 3). This evaluation included two HLA-A*03:01-restricted T cell lines, one HLA-A*11:01-restricted T cell line, two HLA-B*07:02-restricted T cell lines, and one HLA-C*07:02-restricted T cell lines. A complete list of peptides used for temperature-based peptide exchange is provided in Supplemental Table 2. We then compared the staining of T cells by temperature-exchanged MHC-I multimers to those stained with classical folded MHC-I multimers. In the latter case, each MHC-I multimer has to be produced individually, while temperature exchange allows for a batch that can be taken from the freezer and warmed up in the presence of intended peptides before flow analyses. This procedure takes only 1 hr. For this comparison, we generated temperature-exchanged MHC-I multimers for each of the four newly developed alleles and compared their staining performance to that of the classical folded MHC-I multimers at isomass for the same peptide (Figure 4). Additionally, staining intensity was evaluated across multiple dilutions using isomass MHC-I multimer concentrations (Supplemental Figure 4). Overall, the temperature-exchanged MHC-I multimers showed distinct binding to clonal T cells in an antigen-specific manner and exhibited only a minor decrease in performance compared to the classical MHC-I multimers, particularly for the HLA-A11:01 and HLA-B07:02 complexes. Temperature-exchanged multimers for HLA-A*03:01 and HLA-C*07:02 showed similar or slightly lower staining intensities compared to their classical counterparts. Higher concentrations of both multimer types can be used to further enhance staining intensity without any technical complications.

Temperature exchange tetramer binding to clonal T cell lines.

Clonal CD8+ T cell lines stained with PE-conjugated temperature exchange multimers that either contained a relevant (pink) or irrelevant peptide (grey). The specificities of the T cell lines are depicted above each figure, as well as HLA-type and first three amino acids of the peptide. Y-axis shows the individually scaled counts and x-axis shows the mean fluorescent intensity of PE.

Comparison of temperature exchange multimers to conventional multimers on clonal T cell lines.

Clonal CD8+ T cell lines stained with 13.3 ng/μl of PE-conjugated temperature exchange tetramers (pink) or with conventional tetramers (blue). The specificities of the T cell lines are depicted with the HLA-type and first three amino acids of the peptide for each line. Y-axis shows the individually scaled counts and x-axis shows mean fluorescent intensity of PE.

Detection of virus-specific CD8+ T cells in peripheral blood

To evaluate the practical utility of our additional HLA alleles for temperature-based peptide exchange MHC-I multimers using human peripheral blood samples, we compared the performance of temperature-based peptide exchanged multimers to that of classical multimers within an immune monitoring framework. This is particularly relevant in the context of monitoring virus-specific T cells in immunocompromised patients, such as those undergoing T cell-depleted allogeneic stem cell transplantation and adoptive cell therapy. Herpesviruses, including human cytomegalovirus (HCMV) - and Epstein-Barr virus (EBV), can cause severe infections in these patients. Restoring T cells is essential to prevent these infections and to minimize complications and mortality. We analyzed peptides derived from HCMV and EBV for the HLA types HLA-A*11:01, HLA-B*07:02, and HLA-C*07:02 in peripheral blood samples from healthy donors. Additionally, we included the previously developed temperature-based peptide exchange multimers for HLA-A*02:01 in this analysis[18]. We exchanged PE- and APC-labeled multimers for a selection of these peptides to measure T cell frequencies in peripheral blood mononuclear cells (PBMCs) and compared them to traditional folded multimers. The antigen-specific T cells identified with temperature-exchanged multimers stained a distinct cell population (Figure 5). Importantly, we did not observe any MHC-I multimer positive T cells in two irrelevant donors (Supplemental Figure 5). The percentages observed are comparable to those observed with conventional multimers (Figure 5). The peptides are listed in Supplemental Table 2 and the gating strategy of the immune monitoring framework is shown in Supplemental Figure 6. These result illustrates the efficiency and versatility of our technology in rapidly generating diverse MHC-I multimers from a ready-to-use stock, enabling the detection of antigen-specific T cells, even at the low frequencies commonly observed in primary immune monitoring samples.

Comparison of temperature exchange tetramers to conventional tetramers on PBMCs.

PBMCs of healthy individuals were incubated with peptide exchange (pink) or conventional (blue) tetramers. Donor PBMCs were CMV- or EBV-positive based on serology and were incubated with the relevant tetramer based on donor HLA-type. Tetramers were conjugated to APC (y-axis) as well as PE (x-axis). Gate shows the tetramer-binding CD8+ T cells and the percentage of total CD8+ T cells is depicted in the top right corner.

Discussion

The use of multimeric versions of peptide MHC-I complexes has become an important technology to detect and quantify antigen-specific CD8+ T cells[29–32]. To identify antigen-specific T cells using flow cytometry, fluorescently labeled MHC multimers are widely employed. There is an increasing interest in the detection of CD8+ T cells immune responses since they are important for the protection against pathogens or cancer and can be used to define antigens for vaccination purposes. Two challenges in advancing this field are the need for high-throughput approaches and the lack of ready-to-use reagents. Classical peptide-MHC-I complexes are each produced in a separate production run, required for each unique peptide-HLA complex. The process is elaborate, time consuming and demands highly skilled people. This restricts the generation of a larger library of peptide-HLA combinations and thereby results in incomplete detection of the full CD8+ T cell response against an antigen. Previously, it has been shown that peptide-exchange technologies, such as UV-based peptide exchange, are able to upscale towards higher throughput approaches. However, quantifying the efficiency of UV-induced peptide destruction and subsequent peptide exchange remains challenging. Temperature-based peptide exchange is controlled, simple, fast, and quantitative if the correct conditions are identified, as for the HLA alleles in this manuscript. Moreover, the broader availability and simplicity of equipment needed for thermal exchange compared to UV sources enables wider adoption of this approach. The time to prepare the materials for flow analyses is around 1 hour and consists of taking the exchange MHC-I multimer from the fridge and warming it up in the presence of peptides of interest. This can be immediately followed by staining the CD8+ T cells for flow cytometric analysis.

Additionally, MHC-I molecules can interact with other surface proteins in both a peptide-dependent and peptide-independent manner. For example, they bind to the inhibitory receptor Leukocyte immunoglobulin-like receptor subfamily B member 1 (LILRB1), independently of the sequence of the presented peptide, while interactions with receptors such as Killer Immunoglobulin-like Receptors and CD55 are peptide-dependent[33–36]. Consequently, a positive flow cytometry signal does not necessarily indicate interaction with the TCR of CD8+ T cells. It is therefore essential to exercise caution when interpreting MHC-I multimer results to ensure accurate conclusions. These diverse MHC-I based interactions may provide further utility for studying various immune cell subsets beyond T cells, including natural killer (NK) cells and other innate immune cells from a peptide MHC-I specific perspective.

We have set out to determine whether temperature-based peptide-exchange technology can be developed into a broadly applicable platform for the screening of human CD8+ T cells. For this purpose, we designed conditional peptide ligands for four different human MHC-I alleles across the three human HLA loci: HLA-A*03:01, HLA-A*11:01, HLA-B*07:02 and HLA-C*07:02. Refolding reactions with the temperature-based peptide are efficient and result in stable peptide MHC-I complexes at 4°C, which are unstable with increased temperature. The presence of an high-affinity peptide during temperature-based peptide exchange ensures the replacement of the temperature conditional ligand in the MHC-I peptide-binding groove. This leads to the generation of the peptide-MHC-I complex with a desired antigen specificity. Importantly, we show that the temperature-based peptide exchange can be performed using the multimerized and fluorescently labeled MHC-I complexes. This allows the development of ready-to-use MHC-I multimers, with high-throughput abilities demonstrating the ease of use of temperature-based peptide exchange technology.

The temperature-based peptide exchange method for generating MHC-I multimers, has several limitations. First, effective peptide exchange necessitates the design of HLA-specific ligands, adding complexity compared to approaches that utilize empty-loadable MHC-I multimers. This reliance on specific ligands could also affect the specificity and functionality of T cell responses, particularly when low-affinity peptides are involved. Additionally, although the method allows for rapid ready-to-use multimer generation, the inherent stability of some MHC-I complexes could limit the types of peptides that can be successfully exchanged. Furthermore, while this system is designed to be accessible to a wide range of laboratories, the initial need for appropriate peptide selection and optimization might still pose a challenge for some users. This technology is most likely not optimal for low affinity peptides, as the template peptide itself must also have low affinity. Lastly, despite the benefits of temperature-based peptide exchange technology, there is still a risk that some combinations may not provide the desired specificity and functionality. This highlights the need for ongoing refinement and validation of the method across different MHC-I alleles. Additionally, using complementary peptide exchange methods may enhance overall effectiveness as has been done for UV-exchange multimers and empty disulfide-stabilized multimers[37].

Current peptide exchange technologies predominantly target HLA alleles common in individuals of Western European descent. However, we have started expanding this work to include HLA-A*11:01, which is most prevalent in East and Southeast Asian populations [38, 39]. Currently there are temperature-based peptide-exchange multimers for six HLA alleles, namely HLA-A*02:01, HLA-A*03:01, HLA-A*11:01, HLA-B*07:02, HLA-C*07:02 and HLA-E[19, 20]. With this set of MHC-I reagents, a deep analysis could be possible for people with an European genetic background.

A variety of MHC-I reagents are currently available for use in CD8+ T cell diagnostics, including classical folded MHC-I multimers, empty stabilized multimers[9] and UV-exchange multimers. These MHC-I reagents are being used to determine virus specific T cell diagnostics, personalized neoantigen-specific T cell diagnostics and more. We propose that the ready-to-use temperature-based peptide exchange MHC-I technology could further advance the field, as these reagents simplify the process by eliminating additional steps typically required for MHC-I multimers production.

Material and methods

PBMCs isolation

PBMCs were isolated from healthy donors and obtained via Sanquin or biobanked by the Department of Hematology (LUMC) after informed consent, in accordance with the Declaration of Helsinki. PBMCs were isolated from buffy coats or whole blood using Ficoll-Isopaque and used directly or stored using cryo-preservation. Biobanked PBMCs were thawed in Iscove Modified Dulbecco Medium (IMDM; Lonza, Basel, Switzerland) supplemented with 10% heat-inactivated fetal bovine serum (FBS; Sigma-Aldrich, Saint Louis, MO, USA), 2.7 mM L-glutamine (Lonza), 100 U/mL penicillin (Lonza), and 100 µg/mL streptomycin (Lonza; 1% p/s), and subsequently treated with 1.33 mg/mL DNAse to minimize cell clumping. PBMCs were either used for isolation of clonal T cell lines or for flow cytometry analysis.

Clonal T cell lines

Clonal T cell lines were isolated and expanded as previously described[40, 41]. In brief, PBMCs were incubated with peptide-HLA multimers and antibodies targeting CD14, CD8 and CD4 to isolate CD14-CD8+CD4- multimer-binding cells using fluorescence-activated cell sorting (FACS). In some cases, cells were pre-selected prior to FACS using magnetic-activated cell sorting (MACS) based on peptide-HLA multimer-binding. Obtained cells were further expanded either as bulk[40] or single cell per well[41], in the presence of irradiated feeders with or without irradiated Epstein-Barr virus lymphoblastoid cell lines (EBV-LCLs), phytohemagglutinin (PHA) or specific peptides. See Supplemental Table 1 for T cell line details.

Flow cytometry

1-2x106 PBMCs or 0.05-0.1x106 T cells were seeded in a round bottom 96-wells plate. Cells were stained with the viability dye Zombie-Red (Biolegend, San Diego, CA, USA) for 25 min at room temperature after which the cells were washed in PBS containing 0.8 mg/mL albumin (FACS buffer). Cells were stained with 20 μl multimer staining mix, incubated for 30 minutes at 4°C. Conventional multimers were generated in-house, as previously described[40]. After washing, cells were stained with 20 μl of an antibody cocktail containing: αCD8-pacific blue (cat#558207, BD Biosciences), αCD4-BV510 (cat#562970, BD Biosciences), αCD14-FITC (cat#555397, BD Biosciences), αCD19-FITC (cat#555412, BD Biosciences) and αCD16-FITC (cat#335035, BD Biosciences). After incubation for 20 minutes at 4°C, cells were washed and measured on a 3-laser Aurora (Cytek Biosciences) or LSR II (BD Biosciences) flow cytometer. Flow cytometry data analysis was performed using OMIQ.ai. See Supplemental Table 2 for patient details regarding CMV- and EBV status and HLA typing.

K562 LILRB1OE cells were harvested, washed and 100K cells were stained in 50 uL facs buffer (1% FCS in PBS) supplemented with multimers (1:80 dilution for temp exchange or 200 ng conventionel tetramer) or anti-LILRB1 FITC (1:40, clone REA998, miltenyi biotec) for 30 minutes on ice shielded from light. After 2 rounds of washing, cells were measured in 100 uL facs buffer on a BD FACS LSR II 4L Full (BD Biosciences, San Jose, CA, USA) flow cytometer and data was analyzed using FlowJo FlowJo (V10.8.1).

Peptide synthesis and purification

Peptides were synthesized at the LUMC Peptide Facility by standard solid-phase peptide synthesis using Syro I and Syro II synthesizers. Peptides were purified using reversed-phase HPLC (Waters) using a mobile phase of water/acetonitrile gradient containing 0,1% TFA on a C18 column (X-bridge, Waters). Purity of the peptides was confirmed using LC-MS (Micromass LCT Premier; Waters).

Generation of temperature sensitive monomers and multimerization

Temperature-based sensitive monomers were produced as previously described[21]. Briefly, recombinant HLA-A/B/C heavy chain and Β2M light chain were produced as inclusion bodies in bacterial strain BL21 (DE3). These purified inclusion bodies containing 2.5 mg of heavy chain were solubilized in 8 M urea and folded for 5 days at 10°C with 1.2 mg prefolded Β2M and 3 mg temperature-sensitive peptide (A3: RGAVRATKR; A11: RFKMFPEVR; B07: RKELVRPAL; C07: IKYNYPAMA) in 50 ml folding buffer (400 mM L-Arginine, 0,5 mM oxidized glutathione, 5 mM reduced glutathione, 2 mM EDTA, 100 mM Tris.HCl pH 8, glycerol 5% and half a tablet of protease inhibitor cocktail (complete, Roche)). Temperature-based sensitive monomers were concentrated on a 30 kDa filter (Amicon Ultra-15) and biotinylated with BirA ligase overnight at 4°C. Biotinylated Temperature-based sensitive monomers were purified with size exclusion chromatography (Superdex 75 10/300 GL) on a NGC system (Bio-Rad). Properly folded complexes were concentrated on a 30 kDa filter (Amicon Ultra-15) and the concentration was measured on a NanoDrop (Thermo Fisher Scientific) and stored in 300 mM NaCl and 20 mM Tris-HCl, pH 8.0 with 15% glycerol at -80°C. Next, the monomers were multimerized with PE-labeled streptavidin and/or APC-Labeled streptavidin (SA-PE, SA-APC; Thermo Fisher Scientific, Invitrogen) to a final concentration of 0,625 µM. For peptide exchange, temperature-based monomers were incubated at a final concentration of 0,5 µM and 50 µM of selected peptide. This was diluted 1:20 with PBS to be added to CD8+ T cells.

HPLC analysis of temperature-mediated peptide exchange

To initiate temperature-based peptide exchange, a 0.5 µM peptide–MHC-I complex was mixed with 50 µM exchange peptide in 110 µl PBS and incubated under defined conditions. After centrifugation at 14,000 g for 1 min, the supernatant was analyzed using size-exclusion chromatography on a Shimadzu Prominence HPLC system with a 300 × 7.8 mm BioSep SEC–s3000 column and PBS as mobile phase. Data were analysed using Shimadzu LabSolutions software.

LILRB1 over-expressing K562 cell line

The ECD of LILRB1 was fused to the platelet-derived growth factor receptor transmembrane spanning domain and cloned into pCDH (ires puro) vector (deposited at Addgene). Virus production, HEK 293T cells were transfected with packaging plasmids pMDLg/pRRE (Addgene), pRSV-Rev (Addgene), pCMV-VSV-G (Addgene) together with the d_LILRB1 plasmid using polyethyleneimine (Polyscience Inc.) . Virus was harvested, filtered and K562 cells were transduced with LILRB1 in the presence of 8 µg/ml polybrene (Millipore). Transduced cells were selected using puromycin (2 µg/ml, Thermo Fisher) and sorted twice based on high LILRB1 antibody staining (clone REA998, Miltenyi Biotec, Fitc) on a CytoFLEX SRT Benchtop cell sorter (Beckman Coulter, Brea, CA, USA). K562 cells transduced with LILRB1 were cultured in Iscove’s Modified Dulbecco’s Medium (IMDM) supplemented with 10% FBS (Bodinco), 1.5% glutamine (200 mM; Lonza) and 1% penicillin/streptomycin (200 mM; Lonza). The stability of PE fluorescent tetramers was indicated by PE-positive K562 cells[27, 36].